Do researchers preferentially collaborate with same-gendered colleagues?

### 2018-05-23 manuscript/Manuscript_Claire.Rmd
manuscript/

# Discussion
We found evidence that researchers tend to publish more with same-gendered coauthors, despite implementing stringent controls for ???spurious??? homophily resulting from Wahlund effects (\autoref{wahlund_plot}). Our data therefore reaffirm earlier conclusions [@Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] that research collaborations occur between same-gendered colleagues substantially more often than expected if collaborators are selected without regard to their gender. Many journals had $\alpha'$ values over 0.1, meaning that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Few journals had $\alpha'$ values below zero, and fewer still had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered collaborators.

As some disciplines exhibit strong gender-biases (e.g. physics and computer science with predominantly men co-authors, and nursing, midwifery for women [@Holman_2018]), we hypothesized that these disciplines would also display a higher level of gender-homophily due to the increased difficulty to find an opposite-gender co-author. However, we found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines (Figure 2). This finding may be indicative that gender homophily in science is a comprehensive concern, rather than an artifact of the existing gender gap, and not limited to stereotypical disciplines. Furthermore, we found that gender homophily was slightly stronger in research disciplines with a female-biased gender ratio, as compared to male-biased or gender-balanced fields (whether it is in the first or last author position). One interpretation of this result is that men are more likely to seek out other men in fields in which men are in the minority, relative to the homophily shown by women in fields in which women are in the minority.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This result suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with rise in the frequency of same-gender collaborations, not a fall. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the last 10 years [@Bendels_2018; @Long_2015; @McKenzie_2017]. Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff [@Shaw_2012, @West_2013], there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might partially be due to the greater numbers of women working in leading roles.

@@

Why does gender homophily exist in science? Collaboration often leads to co-authorship publications, citations and additional professional recognitions, therefore first author should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a woman colleague working on the same area of expertise was slimmer than now. Even though the number of women in STEM is increasing, and it is not likely to be an obstacle to collaboration nowadays; we cannot forgo that female representation in science is still not equal to that of males. Second, the most basic source of homophily is space, scientists are more likely to collaborate with colleagues closer to them in geographic location (@McPherson_2011). In the case of women, personal considerations (e.g. family commitments) represent the principal restrictions on their international mobility. As a results, women are less likely to originate collaborations with foreign colleagues [@Lemoine_1992; @Lewison_2001; @ Webster_2001; @ Lariviere_2011]. Lastly, men and women think differently on average (REF), and are more likely to initiate collaborations based on similarity of thoughts to progress their scientific output (REF).

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work, and is frequently used as a success indicator for being promoted and hired as a faculty member. Our finding that researchers are more likely to publish with same-gendered coauthors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are therefore less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@West_2013], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. And as a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992].

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more than their male counterpart [@Debarre_2018] . Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events[@Isbell_2012; @Schroeder_2013; @Klein_2017], discrimination during hiring [@MossRacusin_2017; @Reuben_2014], discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], all in all this concentration of small disadvantages drives women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.

Altogether, our findings suggest that researchers preferably collaborate with same-gendered colleagues in STEMM research. Despites using a more stringent analysis methods (e.g. Wahlund effects) and the largest dataset to date, our results were in accord with longstanding and recent research. Gender homophily in science not only reinforces the less favorable status of women in science, but also creates long term problems (see above). It is important to explore the causes and implement measures to redress the gender disparity. We advocate for the introduction and implementation of gender policies in order to support opposite sex collaboration, to give men and women a chance to collaborate across gender, fields, and to promote scientific and innovative excellence.

>>>>

# Discussion
We found evidence for researchers tending to publish more constantly with same-gendered coauthors, despite implementing stringent controls for ???spurious??? homophily, resulting from Wahlund effects (\autoref{wahlund_plot}). Therefore, our data reaffirm earlier conclusions [@Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] that research collaborations between same-gendered colleagues occur substantially more frequently than expected by random, not considering their gender. Many journals had $\alpha'$ values over 0.1, implying that the gender ratio of the average man's coauthors is skewed towards men by >10%, relative to the gender ratio of a woman's coauthors. Only few journals had $\alpha'$ values below zero, and even fewer had significantly negative $\alpha'$ values, suggesting that it is comparatively rare for researchers to preferentially select opposite-gendered co-authors.

As some disciplines exhibit strong gender-biases (e.g. physics and computer science with predominantly male co-authors, and nursing and midwifery female [@Holman_2018]), we hypothesized that these disciplines would also display a higher level of gender-homophily due to the increased difficulty to find an opposite-gender co-author. However, we found no evidence that gender homophily is restricted to particular disciplines: instead, there was evidence of homophily across many diverse research disciplines (Figure 2). This finding may be indicative that gender homophily in science is a comprehensive concern, rather than an artifact of the existing gender gap, and not limited to stereotypical disciplines. Furthermore, we found that gender homophily was slightly stronger in research disciplines dominated by a female-biased gender ratio, compared to male-biased or gender-balanced fields (regardless of their position). One interpretation of this result is that men are more likely to seek out to publish with other men in fields where men are in minority, compared to the homophily displayed by women in fields where females are in in minority.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. This suggests that recent increases in the number of women working in STEMM, as well as recent efforts to recruit, retain, and celebrate women in STEMM, have co-occurred with a rising frequency of same-gender collaborations. We speculate that this trend might result, at least in part, from the increasing number of women working in senior positions in STEMM over the past decade [@Bendels_2018; @Long_2015; @McKenzie_2017]. Because many co-authorships involve a junior and a senior researcher (e.g. a student and professor), and the gender ratio is typically more skewed among senior staff [@Shaw_2012, @West_2013], there will be more opposite-gender collaborations than expected if the gender ratio were the same at both career stages. Thus, the observed increase in homophily might partially be due to the greater number of women working in leading roles.

@@

Why does gender homophily exist in science? Collaboration often leads to co-authorship publications, citations and additional professional recognition, therefore first authors should always be considering collaborators that are most likely to enhance their productivity and in the long term also their academic success, rather than pick co-authors according to their gender. First of all, it is likely that decades ago, the probability to meet a female colleague working on the same area of expertise was slimmer than now. Even though the number of women in STEMM is increasing, i.e. it is not likely to be an obstacle to collaboration anymore; we cannot forgo that female representation in science is still not equal to that of males. Second, the most basic source of homophily is space, scientists are more likely to collaborate with colleagues closer to them in geographic location (@McPherson_2011). In the case of women, personal considerations (e.g. family commitments) represent the principal restrictions on their international mobility. As a result women are less likely to originate collaborations with foreign colleagues [@Lemoine_1992; @Lewison_2001; @ Webster_2001; @ Lariviere_2011]. Lastly, men and women think differently on average (REF) and are more likely to initiate collaborations based on similarity of thoughts to progress their scientific output (REF).

Academic publishing is one of the most (if not the most) important tool for a researcher to display their work and is frequently used as an indicator of success for being promoted and hired as a faculty member. Our finding that researchers are more likely to publish with same-gendered co-authors implies that co-authorship listing is not a random process, but instead creates additional gender imbalance. Homophily could make it harder for women to find collaborators in disciplines where women are in the minority [i.e. almost all STEMM fields; @Holman_2018]. Because women make up for as little as 30% of the STEMM scientist community [@Holman_2018], they are less likely to be picked as coauthors, and, as a result, gender homophily creates further gender imbalance and disadvantages for women. Men tend to publish more in high impact journals as first author [@West_2013], and generate more citations than papers lead by first author women [@Long_1992; @Lariviere_2013]. As a consequence, gender homophily is also likely to contribute to lower publication rates for women [@McDowell_1992].

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more often than their male counterparts [@Debarre_2018]. Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events [@Isbell_2012; @Schroeder_2013; @Klein_2017], discrimination during hiring [@MossRacusin_2017; @Reuben_2014] and discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], are only some of the small disadvantages adding up to drive women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.

Altogether, our findings suggest that researchers preferably collaborate with same-gendered colleagues in STEMM research. Despites using a more stringent analysis method (e.g. Wahlund effects) as well as the largest dataset to date, our results were in accord with longstanding and recent research, while adding new details, reasoning and insight for the current situation. Gender homophily in science not only reinforces the less favorable status of women in science, but also creates long term problems as described above. It is paramount to explore the causes and implement measures to readdress the gender disparity. We advocate for the introduction and implementation of gender policies in order to support opposite sex collaboration, to give men and women a chance to collaborate across gender, fields, and to promote scientific and innovative excellence.
